# Bivariate binormal model {#bivariate-models}


```{r setup, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
  library(RJafroc)
  library(ggplot2)
  library(mvtnorm)
  library(gridExtra)
  library(grid)
  library(png)
```


## How much finished 30% {#bivariate-models-how-much-finished}



## Introduction {#bivariate-models-introduction}

Until now the focus has been on the single rating per case scenario where a reader interprets a set of cases in a single modality. This chapter describes two approaches to the problem of two readers interpreting a common set of cases. Because of the paired interpretation the two ratings per case are correlated. The first approach extends the binormal model to paired interpretations while the second extends the contaminated binormal model to paired interpretations. The corresponding software implementations are CORROC2 and CORCBM, respectively.


In [@chakraborty2017observer] Chapter 21 was devoted to the CORROC2 algorithm and usage of the software was described in detail. At that time [@zhai2017bivariate] was in press - this paper describes an alternate algorithm -- termed the correlated contaminated binormal model or CORCBM -- that has several advantages over CORROC2. Therefore, I have reconsidered the focus of this chapter: it now outlines the correlated binormal model fitting procedure but does not detail how to use it. Instead I now include a detailed explanation of the work presented in [@zhai2017bivariate] and how to use the CORCBM software implementing it. The included description of the CORROC2 model - basically how to extend the binormal model to a correlated binormal model - will help the reader better understand how to extend the contaminated binormal model to the correlated contaminated binormal model. 


In this chapter the bivariate extension of the (univariate) binormal model is described. Bivariate sampling is described at a relatively simple level. Interactive visualization of the probability density functions is demonstrated. Estimation of parameters of the bivariate binormal model from ratings data is addressed. 


## The bivariate binormal model {#bivariate-models-binormal}

Sampling from the bivariate normal distribution $N_2$ is as follows:


\begin{equation}
\overrightarrow{x} \sim  N_2\left( \overrightarrow{\mu}, \Sigma \right)
(\#eq:multivariate-sampling-model)
\end{equation} 


Here $\overrightarrow{x}$ is a length-$2$ vector of the observed z-samples for each case, $\overrightarrow{\mu}$ is a length-$2$ vector containing the means of the bivariate normal distribution and $\Sigma$ is the $2 \times 2$ covariance matrix describing the variances and correlations between the observed samples. 

### The bivariate normal density function {#bivariate-models-binormal-density}

The bivariate normal probability density function $f\left( \overrightarrow{x} ~~ \bigg \rvert ~~\overrightarrow{\mu}, ~~ \Sigma \right)$ is defined by:


\begin{equation} 
f\left( x_1,x_2 ~~ \bigg \rvert ~~ \mu_1, \mu_2,  \sigma_1, \sigma_2, \rho \right)
= \frac{1}{2 \pi \sigma_1 \sigma_2 \sqrt{1-\rho^2}}\exp\left( -\frac{t}{2\left( 1-\rho^2 \right)} \right)
(\#eq:bivariate-models-binormal-density-function)
\end{equation} 


where


\begin{equation} 
t=\frac{\left( x_1-\mu_1 \right)^2}{\sigma_1^2}-\frac{2\rho\left( x_1-\mu_1 \right)\left( x_2-\mu_2 \right)}{\sigma_1 \sigma_2}
+\frac{\left( x_2-\mu_2 \right)^2}{\sigma_2^2}
(\#eq:bivariate-models-binormal-density-function2)
\end{equation} 



For a bivariate distribution the covariance matrix is: 


\begin{equation} 
\Sigma=
\left( \begin{matrix}
\sigma_1^2 & \rho \sigma_1 \sigma_2 \\
\rho \sigma_1 \sigma_2 & 
\sigma_2^2
\end{matrix}
 \right)(\#eq:bivariate-models-binormal-covariance-matrix)
\end{equation} 


In these equations $\mu_1$ and $\sigma_1^2$ is the mean and variance for modality 1, $\mu_2$ and $\sigma_2^2$ is the mean and variance for modality 2 and $\rho$ is the correlation between the two modalities.

The `R` function to evaluate Eqn. \@ref(eq:bivariate-models-binormal-density-function) is `dmvnorm()`, for “density of multivariate normal distribution” available via `R` package `mvtnorm` [@R-mvtnorm]. Its usage is illustrated next for the following parameter values:


\begin{equation} 
\left.\begin{aligned}
\overrightarrow{
\mu}&=
\left( \begin{matrix}
\mu_1 \\
\mu_2
\end{matrix}
\right)
=\left( \begin{matrix}
1.5 \\
2.0
\end{matrix}
\right) \\
\sigma_1 &= 1.1 \\
\sigma_2 &= 1.5 \\
\rho &= 0.6 \\
\Sigma&=
\left( 
\begin{matrix}
\sigma_1^2 & \rho \sigma_1 \sigma_2 \\
\rho \sigma_1 \sigma_2  & \sigma_2^2
\end{matrix}\right)
=
\left( 
\begin{matrix}
1.1^2 & 0.6 \times 1.1 \times 1.5 \\
0.6 \times 1.1 \times 1.5  & 1.5^2
\end{matrix}\right)\end{aligned}\right\}
(\#eq:bivariate-models-binormal-parameters)
\end{equation} 

In the following code the call to get the pdf occurs at line 10.

```{r, attr.source = ".numberLines"}
mu1 <- 1.5 # mean modality 1
mu2 <- 2.0 # mean modality 2
var1 <- 1.1^2  # variance modality 1
var2 <- 1.5^2 # variance modality 2
rho <- 0.6 # correlation between modalities 1 and 2
# construct covariance matrix Sigma
Sigma <- matrix(c(var1, rho*sqrt(var1*var2), 
                  rho*sqrt(var1*var2), var2),2)
x <- c(0.1,0.2) # the x-vector at which to evaluate pdf
pdf  <- dmvnorm(x, mean = c(mu1,mu2), sigma = Sigma)
cat("density at x1 = 0.1 and x2 = 0.2 = ", pdf, "\n")
```


The parameters describing the bivariate normal distribution are $\overrightarrow{\mu}$ and $\Sigma$. The total number of parameters is five: two means, two variances and a correlation coefficient. 



### The bivariate binormal model {#bivariate-models-binormal-formulae}

The bivariate binormal model is an extension of the bivariate model, just described, to *two truth-states*: non-diseased and diseased, each modeled separately as a bivariate normal distribution. This could potentially result in a doubling of the number of parameters but by appropriate shifting and scaling transformations one can ensure that the two means of the bivariate distribution for non-diseased cases are zeroes and the corresponding variances are unity. This reduces the total number of parameters to six: $\mu_1$, $\mu_2$, $\sigma_1$, $\sigma_2$, $\rho_1$, $\rho_2$. Here $\mu_1$, $\mu_2$ are the means for diseased cases for modality 1 and modality 2, respectively, $\sigma_1^2$, $\sigma_2^2$ are the variances for diseased cases for modality 1 and modality 2, respectively, and $\rho_1$, $\rho_2$ are the correlations between the two modalities for non-diseased cases and diseased cases, respectively. 


The bivariate binormal parameters are:


\begin{equation} 
\left.\begin{aligned}
\overrightarrow{
\mu_1}&=
\left( \begin{matrix}
0 \\
0
\end{matrix}
\right)\\
\overrightarrow{
\mu_2}&=
\left( \begin{matrix}
\mu_1 \\
\mu_2
\end{matrix}
\right)\\
\Sigma_1&=
\left( 
\begin{matrix}
1 & \rho_1 \\
\rho_1  & 1
\end{matrix}
\right) \\
\Sigma_2 &= \left( 
\begin{matrix}
\sigma_1^2 & \rho_2 \sigma_1 \sigma_2 \\
\rho_2 \sigma_1 \sigma_2  & \sigma_2^2
\end{matrix}
\right)
\end{aligned}\right\}
(\#eq:bivariate-models-binormal-all-parameters)
\end{equation} 


On the left hand side of the above equation the subscripts refer to the two truth states (1 for non-diseased and 2 for diseased). 


The decision variable is $z_{ik_tt}$ where the $i$ subscript corresponds to the two modalities and the $t$ subscript corresponds to the two truth states. The correlated ratings pairs $\left( z_{1k_11},z_{2k_11} \right)$ and  $\left( z_{1k_22},z_{2k_22} \right)$, corresponding to z-samples from non-diseased and diseased cases, respectively, are abbreviated to $\overrightarrow{z_{1k_tt}}$ (the vectorization always "condenses" the two modalities):
 
 
\begin{equation} 
\overrightarrow{z_{k_tt}}=
\left( \begin{matrix}
z_{1k_tt}  \\
z_{2k_tt}
\end{matrix} \right)
(\#eq:bivariate-models-binormal-notation)
\end{equation} 


<!-- The sampling for each modality is as follows: -->


<!-- \begin{equation} -->
<!-- \left.\begin{aligned} -->
<!-- z_{k_11} &\sim  N\left( 0, 1 \right) \\ -->
<!-- z_{k_22} &\sim  N\left( \mu, \sigma^2 \right) -->
<!-- \end{aligned}\right\} -->
<!-- (\#eq:univariate-binormal-model-sampling) -->
<!-- \end{equation}  -->


<!-- where the parameters $\mu$ and $\sigma$ are of course modality dependent.  -->

$\overrightarrow{z_{k_tt}}$ is sampled from $N_2\left( \overrightarrow{\mu_t}, \Sigma_t \right)$, the bivariate normal distribution with mean $\overrightarrow{\mu_t}$ and covariance matrix $\Sigma_t$:



\begin{equation}
\overrightarrow{z_{k_tt}} \sim  N_2\left( \overrightarrow{\mu_t}, \Sigma_t \right)
(\#eq:bivariate-models-binormal-sampling)
\end{equation} 



The parameters $\overrightarrow{\mu_t}, \Sigma_t$ are truth-state dependent as described in Eqn. \@ref(eq:bivariate-models-binormal-all-parameters). 

<!-- In Eqn. \@ref(eq:bivariate-models-binormal-sampling), the symbol $\sim  N_2\left( \overrightarrow{\mu_t}, \Sigma_t \right)$ denotes a vector sample of length two from the bivariate normal distribution with mean $\overrightarrow{\mu_t}$ and covariance matrix $\Sigma_t$.  -->

<!-- Each case yields two samples corresponding to the two modalities. The subscript 2 on $N$ distinguishes the bivariate distribution from the corresponding univariate distribution $N$ (which has an implicit 1 subscript).  -->

For non-diseased cases the correlation coefficient of the paired samples is $\rho_1$ and the corresponding value for diseased cases is $\rho_2$. With reference to Eqn. \@ref(eq:bivariate-models-binormal-all-parameters) the ones along the diagonal of $\Sigma_1$ confirm that the variances of the individual modality samples for non-diseased cases are each equal to unity. The corresponding variances for diseased cases are $\sigma_1^2$ and $\sigma_2^2$. 


To complete the model one needs to include threshold parameters and a decision rule. For an R-rating single modality ROC task with allowed ratings $r = 1, 2, ..., \text{R}$, one needs $\text{R}-1$ thresholds $\zeta_1,\zeta_2,...,\zeta_{\text{R}-1}$. Defining $\zeta_0 = -\infty$ and $\zeta_{\text{R}} = +\infty$ the decision rule is to label a case with rating $r$ if the realized z-sample satisfies $\zeta_{r-1} < z \le \zeta_r$. 


In the two-modality task the decision variable and the thresholds are modality dependent: i.e., $z_{1k_tt}$ and $z_{2k_tt}$ and two sets of thresholds are needed: $\zeta_{11},\zeta_{12},...,\zeta_{1 (\text{R}-1)}$ and $\zeta_{21},\zeta_{22},...,\zeta_{2 (\text{R}-1)}$. As before $\zeta_{10} = -\infty$ and $\zeta_{20} = -\infty$ and $\zeta_{1\text{R}} = +\infty$ and $\zeta_{2\text{R}} = +\infty$. The decision rule is to rate a case in modality $1$ with rating $r$ if $\zeta_{1(r-1)} < z_{1k_tt} \le \zeta_{1r}$ and the same case in modality $2$ is rated $s$ if $\zeta_{2(s-1)} < z_{2k_tt} \le \zeta_{2s}$.
 	
 	
The sampling model Eqn. \@ref(eq:bivariate-models-binormal-sampling) is a natural extension to two modalities of the single modality univariate binormal model described in Chapter \@ref(binormal-model): 

>The univariate model yields one z-sample per case: samples for non-diseased cases have mean zero and unit variance; samples for diseased cases have mean $\mu$ and variance $\sigma^2$. In contrast the bivariate model yields two z-samples per case. For non-diseased cases the samples have zero mean and unit variance and correlation $\rho_1$. For diseased cases, the samples for the first modality have mean $\mu_1$ and variance $\sigma_1^2$, for the second modality they have mean $\mu_2$ and variance $\sigma_2^2$ and their correlation is $\rho_2$. 



### Visualizing the bivariate density function {#bivariate-models-binormal-multivariate-density-visualization}

It is helpful to visualize the pdfs. Since one is dealing with a vector $\overrightarrow{x}$ of length 2, one needs two axes to depict $x_1$ and $x_2$ and a third axis to depict the pdf. The R-package `plotly` [@R-plotly] provides a visualization technique. It is interactive: by dragging the cursor over the plot one can visualize it from different angles. Shown in Fig. \@ref(fig:bivariate-models-binormal-pdf-plots) are four views of the bivariate distribution density function for $\mu_1 = 1.5$, $\mu_2 = 2.0$, $\sigma_1 = 1.0$, $\sigma_2 = 1.5$, $\rho_1 = 0.3$ and $\rho_2 = 0.6$.



<!-- ```{r, echo=FALSE} -->
<!-- mu1 <- 1.5 # diseased mean modality 1 -->
<!-- mu2 <- 2.0 # diseased mean modality 2 -->
<!-- var1 <- 1  # diseased variance modality 1 -->
<!-- var2 <- (1.5)^2 # diseased variance modality 2 -->
<!-- rho1 <- 0.3 # non-diseased correlation -->
<!-- rho2 <- 0.6 # diseased correlation -->

<!-- # -->
<!-- # construction the covariance matrix Sigma1 for non-diseased -->
<!-- #  -->
<!-- Sigma1 <- matrix(c(1, rho1, rho1, var2),2) -->

<!-- x <- c(0.1,0.2) # the x-vector at which to evaluate pdf -->

<!-- # -->
<!-- # pdf for non-diseased -->
<!-- #  -->
<!-- # following defaults to zero mean for each modality -->
<!-- pdf1  <- dmvnorm(x, sigma = Sigma1)  -->

<!-- # -->
<!-- # construction the covariance matrix Sigma2 for diseased -->
<!-- #  -->
<!-- Sigma2 <- matrix(c(var1, rho2*sqrt(var1*var2), rho2*sqrt(var1*var2), var2),2) -->

<!-- # -->
<!-- # pdf for diseased -->
<!-- #  -->
<!-- pdf2  <- dmvnorm(x, mean = c(mu1,mu2), sigma = Sigma2) -->

<!-- # -->
<!-- # print results -->
<!-- #  -->
<!-- cat("mu1 =", mu1, ", mu2 =", mu2, "\n") -->
<!-- cat("var1 =", var1, ", var2 =", var2, "\n") -->
<!-- cat("rho1 =", rho1, ", rho2 =", rho2, "\n") -->
<!-- cat("non-diseased density at x1 = 0.1 and x2 = 0.2 = ", pdf1, "\n") -->
<!-- cat("diseased density at x1 = 0.1 and x2 = 0.2 = ", pdf2, "\n") -->
<!-- ``` -->


```{r echo=FALSE, bivariate-models-binormal-pdf-plots, fig.cap="Bivariate pdf plots for the bivariate distribution density function for $\\mu_1 = 1.5$, $\\mu_2 = 2.0$, $\\sigma_1 = 1.0$, $\\sigma_2 = 1.5$, $\\rho_1 = 0.3$ and $\\rho_2 = 0.6$.. Top-left is the default view. Top-right is the 'birds eye' view: the lower-left blob is the diseased distribution. Bottom-left is the view looking along the y-axis: the left distribution is non-diseased. Bottom-right is the view looking along the x-axis: the higher peak distribution is non-diseased.", fig.show='hold'}
p1 <-  readPNG("images/bivariate/Fig1A.png")
p2 <-  readPNG("images/bivariate/Fig1B.png") 
p3 <-  readPNG("images/bivariate/Fig1C.png") 
p4 <-  readPNG("images/bivariate/Fig1D.png") 
grid.arrange(grobs = list(rasterGrob(p1),rasterGrob(p2),rasterGrob(p3),rasterGrob(p4)),
              nrow =2,ncol=2)
```



### Estimating bivariate binormal model parameters {#bivariate-models-binormal-multivariate-density-estimation}

In Chapter \@ref(binormal-model) a method for estimating the parameters of the univariate binormal model was described. The method involved maximizing the likelihood function, i.e., the probability of the observed data as a function of the model parameters. The likelihood function was maximized with respect to these parameters. The values of the parameters at the maximum are the maximum-likelihood estimates (MLEs).

With a bivariate model one is dealing with six non-threshold parameters $\overrightarrow{\mu_t}, \Sigma_t$ plus threshold parameters for each modality. Again, the starting point is the likelihood function, i.e., the probability of the observed data as a function of the parameter values. The non-diseased counts in bin $r$ of the first modality and bin $s$ of the second modality is denoted by $K_{rs1}$ (i.e., the binning indices occur before the truth index). The corresponding diseased counts are denoted  $K_{rs2}$. Each case yields two integer ratings, $r$ and $s$. To construct the matrix $K_{rs1}$ one starts with a zero-initialized R x R matrix, then one increments the cell at row $r$ and column $s$ by unity for each non-diseased case that received a $r$ rating in the first modality and an $s$ rating in the second modality. The procedure is repeated for the diseased cases yielding $K_{rs2}$.

For non-diseased cases, the probability of a z-sample in bin r of the first modality and bin s of the second modality is determined by the “volume” under the bivariate distribution $N_2(\overrightarrow {\mu_1}, \Sigma_1)$ between modality-1 thresholds $\zeta_{1(r-1)}$ and $\zeta_{1r}$, and between modality-2 thresholds $\zeta_{2(s-1)}$ and $\zeta_{2s}$. For diseased cases the corresponding probability is the volume under the bivariate distribution $N_2(\overrightarrow {\mu_2}, \Sigma_2)$ between the same thresholds. The probabilities can be calculated using the `pmvnorm()` function, with appropriate parameters, in R package [@R-mvtnorm].    

The probability of observing $K_{rs1}$ non-diseased and $K_{rs2}$ diseased counts in bin $r$ in the first modality and bin $s$ in the second modality is:
 

\begin{equation} 
\prod_{t=1}^{2}\left ( p_{rst} \right )^{K_{rst}}
\end{equation} 


The logarithm of the likelihood function is given by (neglecting a combinatorial factor that does not depend on the parameters):

\begin{equation} 
LL\left ( \mu_1, \mu_2, \sigma_1, \sigma_2, \rho_1, \rho_2, \overrightarrow{\zeta_1},\overrightarrow{\zeta_2} \right ) = \sum_{t=1}^{2}\sum_{s=1}^{R}\sum_{r=1}^{R} K_{rst} \log(p_{rst})
\end{equation} 


The maximum likelihood estimates of the parameters are obtained by maximizing the LL function [@metz1980statistical, @metz1984new] which is implemented in CORROC2. The software measures ratings-correlations at the underlying z-sample level. Much as I have emphasized that ratings are not "hard" numbers the CORROC2 estimated correlations are valid because the algorithm models the ratings as continuous variables and estimates the correlation based on the bivariate binormal model.

## Comments on the bivariate binormal model

CORROC2 is one of the relatively under-utilized tools developed by Prof. Charles E. Metz. There are only four publications describing it TBA 4-6,8 and two of them are difficult to find5,8, nor has the software been maintained on a level comparable to ROCFIT. For example, a method for assessing the goodness of the fit is currently not implemented. With $R^2$ cells it is almost impossible to maintain at least five counts per cell. 

One reason for the relative under-utilization of CORROC2 could be that it is not designed to analyze multiple readers interpreting a common set of cases in two or more modalities i.e., MRMC datasets. For example, in a study in which CORROC2 was used9, four readers interpreted cases in four modalities. With four modalities, there are six possible combinations (4x3/2), so a conservative Bonferroni type correction for the p-value would be to divide 0.05 by 6. For each reader CORROC2 yields a p-value for the difference between the chosen pairs of modalities. If all readers agree, then there is an unambiguous answer, but the method does not allow for correlations between different readers, which undoubtedly leads to loss of power. More fundamentally, it does not address the question of interest: when averaged over all four readers, are the figures of merit in the four modalities different. 

It is possible to analyze multiple reader multiple modality MRMC-ROC datasets without actually using CORROC2. In chapters Chapter 09 and Chapter 10, it was shown how one analyzes such datasets using the empirical AUCs. Alternatively, one could estimate AUC using the univariate binormal model software (e.g., ROCFIT, or more modern software). The correlations in the resulting AUC values – i.e., FOM  correlations as distinct from ratings correlations - are explicitly accounted for in the significance testing procedure, i.e., the procedure it is not concerned with ratings-level correlations. 

The biggest weakness of CORROC2 is its dependence on the underlying binormal model which, as we have seen, is susceptible to improper ROC fits and degeneracy problems. 

 

CORROC2 was developed ca. 1980 by Metz, Wang and Kronman 5,10. Subsequent revisions to the program were made by Jong-Her Shen and more recently by Benjamin Herman6,7. Its impact cannot be overstated. There are well in excess of 116 citations to this software. One reason is that, at that time, it was the only software allowing analysis of paired datasets. However, no advances have been made in the intervening 3 decades, which would allow fitting, for example, proper ROC curves to paired - and possibly degenerate - datasets. Part of the reason for this neglect is the shift in emphasis to empirical AUC based analysis, which does not require parametric modeling or curve fitting – one simply calculates the trapezoidal AUC. However, empirical AUC based analysis has its own limitations, and pursuing an improvement over the parametric approach that removes the current limitations of CORROC2 would, in the author's judgment, quite apart from the specific interest in developing realistic simulators, be of scientific interest.

As a final note, at the time of writing, 4/30/17, an extension of CORROC2 has been published1 [@zhai2017bivariate]. It is called CORCBM, for correlated CBM. Details are in a document CORCBM.pdf in the online supplementary material directory corresponding to this chapter. The title of the publication is: "A bivariate contaminated binormal model for robust fitting of proper ROC curves to a pair of correlated, possibly degenerate, ROC datasets". It replaces the bivariate binormal model with the bivariate contaminated binormal model; hence its name CORCBM, for correlated CBM. Since CBM was designed to fit practically any single dataset, including degenerate ones, CORCBM is likewise able to fit practically any paired dataset. An application of CORCBM to calibrating a simulator to a data set containing single-modality multiple-reader ratings is described in Chapter 23.

## The bivariate contaminated binormal model {#bivariate-models-contaminated-binormal-density}

TBA

### The bivariate contaminated binormal model {#bivariate-models-contaminated-binormal} 

The bivariate contaminated binormal model is the bivariate extension of the univariate contaminated binormal model described in Chapter \@ref(proper-roc-models). For non-diseased cases the sampling is identical to that of the correlated binormal model (i.e., that used in CORROC2), see Section \@ref(bivariate-models-binormal). The pdf is given by:
TBA

## Discussion / Summary {#bivariate-models-binormal-corroc2-discussion}

